\plaintitle

### 2016-09-20 Paper/proceedings.tex
Figureless version.

@@

@@

In all experiments, participants entered in information using a slider without tick marks, in order to avoid anchoring effects \cite{matejka2016effect}. Moving the slider adjusts a purple trend line by adjusting its slope (in experiments 1 and 3) or its y-intercept (in experiment 2). After moving the slider, participants had to confirm their choice of trend line. The central measure is accuracy, or the difference between the selected slope and the slope of the series if the residuals were removed. We chose this design over a standard binary forced choice design for its greater expressiveness, as well as for the interactive feedback of adjusting the fit by hand. We avoided more expressive designs (such as freeform drawing or other elicitation techniques) in order to constrain the model-space of participants, and to reduce the impact of factors such as accuracy in motor movements and heterogeneity in input devices.

@@

We designed our first experiment to examine how accurate participants were are estimating the direction (slope, amplitude, or curvature) of trends in bivariate visualizations. We examined a variety of common bivariate visualizations. In addition to linear trends, we examined more complex relationships such as quadratic and trignometric trends.

@@

Participants saw one of each combination of 3 chart types (scatterplot, line chart, or area chart), 8 possible slopes $\beta = \pm \{0.1,0.2,0.4,0.8\}$, 4 bandwidths of Gaussian residuals $\sigma = \{0.05,0.1,0.15,0.2\}$, for a total of 96 stimuli. Type of trend (linear, quadratic, or trignometric) was a random factor, with 32 stimuli of each factor level.

@@

We had three hypotheses for the first experiment:
\begin{enumerate}
	\item \textbf{As the bandwidth of the residuals increased, accuracy would decrease}. Increasing the bandwidth of the residuals results in a lower correlation coefficient and higher perceived noise in the bivariate data. Prior work indicates that these related measures correspond to decreased accuracy for aggregate tasks in bivariate visualizations \cite{albers2014task,harrison2014ranking}. 
	\item \textbf{More complex relationships would result in lower accuracy}. 
\end{enumerate}
\subsection{Results}

@@

The plots had the same factor levels as the previous experiment, for a total of 96 stimuli per participant. The uniform offset was an additional random factor for each stimulus.

@@

We had one hypothesis for the second experiment:
\begin{enumerate}
	\item \textbf{Area charts would be subject to within the bar bias.} That is, participants would estimate lower values of y-intercepts of trends than with line charts and scatterplots. 
\end{enumerate}
\subsection{Results}

\section{3. Estimation of Slope is Sensitive to Outliers}

OLS regression operates under the assumption that there is a unimodal, symmetric distribution of residuals surrounding the line of best fit. Extreme outliers violate this assumption, and can result in trend lines that are substantially different from those produced by more robust methods. Visual inspection can accurately identify certain classes of outliers \cite{albers2014task}. Therefore, regression by eye may afford the estimation of both outlier-robust and outlier-sensitive trends. However, cognitive biases such as anchoring (the tendency to overweight the first set of information), availability (the tendency to overweight more recent or extreme information), and the hot-hand fallacy (the tendency to assume that trends of high or low values with continue) can impact how viewers categorize and utilize outliers in their estimations. These biases can impact forecasting tasks~\cite{campbell2009anchoring,ji2001culture}. That is, regression by eye may not uniformly weight outliers, depending on their position in the plot.

@@

Similar to the previous experiments, participants saw one of each of the three graph types, slope, and Gaussian residual bandwidth. The four outlier quantities $\{0,5,10,15\}$ was an additional factor. To maintain a manageable number of stimuli (piloting showed evidence of fatigue for more than 100 stimuli per trial), both the sign of the trend line (positive or negative), and the location of the outliers (beginning, first third, or end) were random factors, with each level apportioned to half and one third of the stimuli respectively. This resulted in 96 total stimuli, in line with prior experiments.

@@

\section{Discussion}
\subsection{Limitations \& Future Work}
\subsection{Conclusion}
\section{Acknowledgments}

>>>>

\input{./figures/figures.tex}

@@

\expFig

@@

In all experiments, participants entered in information using a slider without tick marks, in order to avoid anchoring effects \cite{matejka2016effect}. Moving the slider adjusts a purple trend line by adjusting its slope (in experiments 1 and 3) or its y-intercept (in experiment 2). Fig. \ref{fig:experiment} shows an example experimental task. After moving the slider, participants had to confirm their choice of trend line. The central measure is accuracy, or the difference between the selected slope and the slope of the series if the residuals were removed. We chose this design over a standard binary forced choice design for its greater expressiveness, as well as for the interactive feedback of adjusting the fit by hand. We avoided more expressive designs (such as freeform drawing or other elicitation techniques) in order to constrain the model-space of participants, and to reduce the impact of factors such as accuracy in motor movements and heterogeneity in input devices.

@@

We designed our first experiment to examine how accurate participants were are estimating the direction (slope, amplitude, or curvature) of trends in bivariate visualizations. We examined a variety of common bivariate visualizations. In addition to linear trends, we examined more complex relationships such as quadratic and trigonometric trends.

@@

Participants saw one of each combination of 3 chart types (scatterplot, line chart, or area chart), 8 possible slopes $\beta = \pm \{0.1,0.2,0.4,0.8\}$, 4 bandwidths of Gaussian residuals $\sigma = \{0.05,0.1,0.15,0.2\}$, for a total of 96 stimuli, with an additional 4 validation stimuli otherwise excluded from analysis. Type of trend (linear, quadratic, or trigonometric) was a random factor, with 32 stimuli of each factor level.

@@

We had three hypotheses for the first experiment:
\begin{enumerate}
	\item \textbf{As the bandwidth of the residuals increased, accuracy would decrease}. Increasing the bandwidth of the residuals results in a lower correlation coefficient and higher perceived noise in the bivariate data. Prior work indicates that these related measures correspond to decreased accuracy for aggregate tasks in bivariate visualizations \cite{albers2014task,harrison2014ranking}. 
	\item \textbf{More complex relationships would result in lower accuracy}. Quadratic and trigonometric relationships are visually more complex than linear relationships, and often require more complex statistical methods to analyze. We anticipated that estimation of these less familiar relationships would therefore be more difficult than in the linear case.
	\item \textbf{Estimations would be unbiased}. That is, there would be no systematic over- or underestimation of trends. 
\end{enumerate}
\subsection{Results}
We performed a three-way analysis of covariance (ANCOVA) of the effect of residual bandwidth, graph type and trend type on unsigned error in estimation of trend lines. We included participant ID and the actual slope of the trend line as covariates.

Our results support our first hypothesis: \textbf{larger residuals reduce accuracy at regression by eye}. Bandwidth of the Gaussian used to generate residuals was a significant main effect ($F(1,4554)=950$), and the trimmed mean of absolute error increased monotonically with this bandwidth, from $7\%$ of the actual slope of the trend line when the bandwidth was $0.05$, to $32\%$ when the bandwidth was $0.20$. Figure \ref{fig:sigma} illustrates these results.

Our results fail to support our second hypothesis: \textbf{there was no statistically significant difference in estimation accuracy among linear, quadratic, or trigonometric trends}. Fit type was only a marginal main effect ($F(2,4554)=2.60$, $p=0.074$), and a post-hoc Tukey's test of Honest Significant Difference (HSD) did not identify any significant pairwise-interactions. The trimmed mean of absolute estimation error of linear fits was $17\%$ of the actual slope, in the middle of the $16\%$ error for quadratic fits, and $18\%$ for trignometric fits, providing evidence that the relative unfamiliarity of non-linear trends does not have a significant (in the ordinary sense) impact on performance.

Our results support our third hypothesis: \textbf{there was no statistically significant bias in estimations}. Participants saw an equal number of stimuli with positive and negative trend lines. If estimates of these trends were unbiased, we would expect the average signed error to be close to zero. The average signed error was $0.0008$, far less than the fidelity of the slider used to input guesses (which had $\Delta=0.01$). A Student's T test failed to support ($p=0.70$) the hypothesis that $mu_{error}\ne0$.

@@

The plots had the same factor levels as the previous experiment, for a total of 96 stimuli per participant, with an additional 4 validation stimuli otherwise excluded from analysis. The uniform offset was an additional random factor for each stimulus.

@@

We had one hypothesis for the second experiment:
\begin{enumerate}
	\item \textbf{Area charts would be subject to within the bar bias.} That is, participants would estimate lower values of y-intercepts of trends than with line charts and scatterplots. 
\end{enumerate}
\subsection{Results}
We performed a one-way ANCOVA of the effect of graph type on \emph{signed} error. We included residual bandwidth as a covariate, and participant ID as a random factor.

Our results support our first hypothesis: \textbf{participants systematically overestimated the intercept of trends in area charts, but not in scatterplots or linegraphs}. Graph type was a significant main effect on signed error ($F(2,2431)=27$, $p<0.001$). A post-hoc Tukey's HSD confirmed significant differences in error between the area chart and the other two chart types, but not between scatterplots and linegraphs. The trimmed mean of the signed error of estimations made with area charts was an underestimation of $-0.02$, compared to a trimmed mean of $0.002$ for the other two conditions. This underestimation corresponded to unsigned errors more than twice as large in area charts (trimmed mean of $0.04$ for intercepts that ranged from $\{-0.25,0.25\}$, compared to $0.02$ for linegraphs and scatterplots). Figure \ref{fig:outlier} illustrates this result.

\section{3. Estimation of Slope is Robust to Outliers}

\outlierFig
OLS regression operates under the assumption that there is a unimodal, symmetric distribution of residuals surrounding the line of best fit. Extreme outliers violate this assumption, and can result in trend lines that are substantially different from those produced by more robust methods. Visual inspection can accurately identify certain classes of outliers \cite{albers2014task}. Therefore, regression by eye may afford the estimation of both outlier-robust and outlier-sensitive trends. However, cognitive biases such as anchoring (the tendency to overweight the first set of information), availability (the tendency to overweight more recent or extreme information), and the hot-hand fallacy (the tendency to assume that trends of high or low values with continue) can impact how viewers categorize and utilize outliers in their estimations. These biases can impact forecasting tasks~\cite{campbell2009anchoring,ji2001culture}. That is, regression by eye may not uniformly weight outliers, depending on their position in the plot.

@@

Similar to the previous experiments, participants saw one of each of the three graph types, slope, and Gaussian residual bandwidth. The four outlier quantities $\{0,5,10,15\}$ was an additional factor. To maintain a manageable number of stimuli (piloting showed evidence of fatigue for more than 100 stimuli per trial), both the sign of the trend line (positive or negative), and the location of the outliers (beginning, first third, or end) were random factors, with each level apportioned to half and one third of the stimuli respectively. This resulted in 96 total stimuli, with an additional 4 validation stimuli otherwise excluded from analysis, in line with prior experiments.

@@

Our results support our first hypothesis. \textbf{Participant estimates were closer to the robust trend line than the trend line that included outliers}. Excluding conditions with zero outliers (and so the two trend lines would be identical), the trimmed mean of unsigned error was over $4$ times higher when calculated as a comparison to the OLS trend line ($0.36$) as opposed to the robust trend line ($0.08$). A Student's T-test confirmed that the error as defined by the OLS line was significantly higher than the robust line ($p<0.001$). However, a Student's T test found a significant difference in unsigned error when compared to the robust trend line between estimates where no outliers were present, and those with any amount of outliers ($M=0.07$ vs. $M=0.08$, $p<0.001$). This indicates that participants are not entirely ignoring outliers, although the small effect size suggestes that participants are giving outliers significantly less weight than OLS regression.

%There is a confound between the number of outliers and the OLS trend line: since all outliers had approximately the same magnitude, more outliers create more extreme slopes with respect to the robust trend line, and so (if a participant were strictly ignoring outliers), larger differences in unsigned error. Similarly, outliers at the ends of the data series will adjust the slope of the OLS line more than outliers in the middle. We therefore measured performance as the unsigned error as a percentage of the difference between the slopes of the two types of fit. We performed a two-way ANOVA on the effect of outlier number and location on this percentage error measure. Participant id was included as a random variable.

%Our results fail to support our second hypothesis. \emph{Increasing number of outliers does not increase fidelity to the OLS trend line}. A one-way ANCOVA on the effect of outlier numbers greater than 0 on this percentage error showed that outliers were a significant main effect ($F(1,3438)=59$, $p<0.001$). However, the percentage error \emph{increases} as the number of outliers increase. A post-hoc Tukey's HSD shows that the percentage errors for 10 and 15 outliers are similar (trimmed means are $111\%$ and $112\%$ respectively), they are significantly different from the stimuli with only 5 outliers (trimmed mean of $97\%$).

\section{Discussion}

In many cases, designers do not explicitly encoded regression information. In other cases, viewers may not have the statistical or graphical expertise to interpret such information, even when it is supplied. Yet, our results point to regression by eye as a robust and accurate method for estimating trends in bivariate data. Participants from a variety of backgrounds and levels of self-reported graphical and statistical expertise were capable of estimating both the slope and intercept of both linear and non-linear trends. That is, viewers of visualization are largely trustworthy when estimating the relationship between two variables in a plot.

However, this general accuracy at estimation trends is not universal. Area charts are visually asymmetrical: area below the line is filled in, and the area above it as not. This asymmetry creates within the bar bias: a systematic underestimation of the intercept of trends, due to the perceived higher likelihood of points in the filled-in area. Designers hoping to rely on regression by eye should avoid such asymmetries in their bivariate visualizations.

Likewise, viewers do not give the same weight to outliers as certain standard forms of statistical regression such as OLS. For noisy data, this robustness may be beneficial: humans themselves can be relied on to perform filtering operations without explicit guidance. However, in other cases, this insensitivity may result in visual estimations at odds with the statistical modeling underpinning other areas of analysis of a particular dataset. The emergence of new modes and the shifting of means may not result in a subsequent shift in beliefs about trends in data.

\subsection{Limitations \& Future Work}

Our experimental setup was intentionally simplistic, affording only a single free parameter for each experiment. In actual regression by eye, the viewer must simultaneously engage in multiple types of estimation: choosing a particular type of fit, filtering data, and then estimating the parameters of the chosen model. Errors in any one of these steps could compound, resulting in performance worse than our measures, where many of these decisions are presented as a given.

There are also a number of design decisions not considered in this study that could impact regression by eye. For example, Wood et al.~\cite{wood2012sketchy} have shown that ``sketchiness'' in visualizations can result in skepticism of the data and design. It is possible that outliers could receive less visual (and so statistical) weight as a response to this sort of skepticism. Our stimuli likewise contained a sufficient number of points that bar charts were not a feasible choice of visualization. Given the propensity of bar charts to encourage comparison of individual, rather than aggregate quantities~\cite{zacks1999bars}, it is possible that bar charts of smaller scale bivariate data could promote fitting of local rather than global trends.

Finally, we focused on a simple form of regression, OLS, as our gold standard for measuring accuracy. While our data were constructed to satisfy the assumptions of OLS (with the exception of our outlier experiment), in most real world scenarios OLS is just one tool of many, and analysts must exercise judgment when determining how to fit their data. More complex forms of regression, and more complex fits, may not have ready analogues visually, and data concerning regression by eye may not extend to cover these cases.

Our future work is focused in three areas. Firstly, we wish to examine the impact of annotations relevant to regression (such as confidence bands, error bars, and curve boxplots~\cite{mirzargar2014curve}) on regression by eye. Can sufficient information promote caution in judgments of trend, or do initial visual estimations have too much weight to be overcome? Secondly, we wish to examine the impact of different rhetorical framings and presentations on regression by eye. Language from semiotics and rhetoric can provide testable structures for how visualizations are consumed~\cite{hullman2011visualization}: these framings (and so associated predispositions or biases about the data) could similarly impact statistical judgments. Lastly, we wish to examine techniques for overcoming bias in regression by eye and similar visual estimations of statistical quantities. Cognitive and perceptual biases are difficult to overcome, and may require exploration of novel forms of visualization \cite{micallef2012assessing}.

\subsection{Conclusion}

In this paper, we examine the ability of viewers of visualizations to perform regression by eye --- the visual estimation of trend in bivariate visualizations. We show that even viewers without statistical training can reliably estimate even non-linear trends in visualizations like scatterplots and linegraphs. However, area charts are subject to within the bar bias, and promote trends with lower intercepts than other bivariate visualizations. Regression by eye is also less sensitive to outliers than simple statistical regression techniques, resulting in a divergence between trends estimated by viewers, which do not give much weight to extreme outliers, and those trends calculated by simple statistical methods, can give equal weight all points in a dataset.

\section{Acknowledgments}
